Symmetric Circuits for Rank Gates

### 2017-08-21 sections/symmetric-circuits.tex
I have realised that my generalisation of symmetry is flawed and will need an overhaul. However, I want to keep a copy so I'm updating git.
This version is the told one where we are just speaking of $\mathcal{D}$-invarient functions.

\documentclass[../paper.tex]{subfiles}
%\usepackage{../mymacros}
\begin{document}
% Note: include circuits in background
Circuits are a well-studied model in complexity theory, and more recently they
have been studied in the context of finite model theory. However, there are some
subtleties in the definition and evaluation of circuits which must be considered
carefully and have important implications for the notion of a symmetric circuit
studied in this paper. Let $C$ be a Boolean circuit $g$ be an internal gate
labelled by some function $f_g: \{0,1\}^{A} \rightarrow \{0,1\}$, for some index
set $A$ with $\vert A \vert = \vert H_g \vert$. Formally, evaluating $g$
requires a (usually bijective) indexing function $f_i: A \mapsto H_g$, which
organises the inputs of $g$ into an appropriate sequence. The evaluation of $g$
is then given by $C[\vec{x}](g) = f_g (a \mapsto C[\vec{x}](f_i(a)))$. As such
the evaluation of a gate in a circuit may depend on how the inputs to the gate
are indexed.

In most contexts in complexity theory, circuits may be taken to be implicitly
ordered objects and so this indexing function can be defined as taking the $i$th
element of $A$ to the $i$th element of $H_g$. However, in general this renders
the evaluation of a circuit for a given input a property both of the circuit and
an arbitrary order. In order to address this problem this problem many authors
only define circuits over Boolean bases of symmetric functions (e.g. the
standard basis or the majority basis). In this case the evaluation of a gate is
independent of the choice of indexing function.

However, in this paper we require gates labelled by Boolean rank functions, i.e.
a function that takes in a Boolean sequence indexed by `row' and `column' sets,
outputs 1 iff that matrix, when interpreted as having entries in $\mathbb{F}_p$,
has rank at least $r$. These rank functions clearly depend on how the input
sequence is labelled, and so are not symmetric. However, it is worth noting that
while such rank functions may be symmetric, they are symmetric in the weaker
sense of being constant under permutations on the rows and columns of the input
matrix.

Motivated by the desire to study circuits with gates that compute rank, We
develop a general framework for circuits that allows gates to be labelled by
Boolean functions of the form $F: \{0,1\}^{X} \rightarrow \{0, 1\}$, where the
labelling $X$ is the universe of a many-sorted structure over some vocabulary
$\tau$. The Moreover, we generalise the notion of a circuit for structures (see
Anderson and Dawar \cite{AndersonD17}), incorporate an appropriate labelling for
the inputs of each gate and allowing for a Boolean basis including non-symmetric
functions. Furthermore, the vocabulary $\tau$ will allow us to impose natural
symmetry conditions on $F$, and so allowing us to develop a useful notion of a
symmetric circuit.

We also prove that no family symmetric functions defined over a basis of
symmetric functions can improve on the expressive power of the circuit model of
Anderson and Dawar \cite{AndersonD17}. Together with the main theorem of this
paper this implies that these more general symmetric circuits are strictly more
powerful.

@@

\subsection{Structured Functions and Symmetry}
Let $\tau = (R, S, \nu)$ be a many-sorted, relational signature, with $R = \{R_1
\ldots, R_t\}$ and $S = [s]$ for some $s \in \nats$, and $\nu(R_i) = (a^i_1,
\ldots , a^i_{r_i)})$, where $r_i = \arty(R_i)$ and $a^i_q \in S$ for all $q \in
[r_i]$. For each $s \in S$ let $A_s$ be a non-empty set, and let $A := A_{1}
\sqcup \ldots \sqcup A_{s} = \{(a,s) : a \in A_s\}$. Let
\[
  \ind (A, \tau) := A_{a^1_1} \times \ldots \times A_{a^1_{r_1}} \times \ldots
  \times A_{a^t_1} \times A_{a^t_{r_t}},
\]. We call $\ind (A, \tau)$ the \emph{index defined by} $(A, \tau)$.

% It is worth noting that there is a natural bijection $P$ that maps
% $\tau$-structures with universe $A$ to subsets of $\Ind(A, \tau)$, taking each
% $\tau$-structure $\mathcal{A}$ to the set $\{(\vec{X}_1, \ldots , \vec{X}_t) :
% \exists i \in [t] \(\vec{X_i} \in R^{\mathcal{A}}_i \) \}$.

For the remainder of the section let $X := \ind (A, \tau)$. Let $G \leq
\sym_{X}$, there is a natural action of $G$ on $X$. There is also a natural
action of $\sym_{A_1} \times \ldots \times \sym_{A_s}$ on $X$ and a subgroup
$G'$ of $\sym_X$ corresponding to the same action. We identify these two groups.

Furthermore, we can defined an action of $G$ on functions of the form $f: X
\rightarrow H$, for any set $H$, given by $\sigma \cdot f(x) = f(\sigma \cdot
x)$ for all $x \in X$. If $H = \{0,1\}$ we say that $f$ is a \emph{binary
  sequence indexed by $X$}. If $H$ is surjective, we say that \emph{$f$ indexes
  $H$ by $X$} or \emph{$f$ indexes $H$ by $(A,\tau)$}.

We develop a notion of equivalence from this action. Say that two functions $f,
g : X \rightarrow H$ are \emph{$G$-equivalent} iff there exists $\sigma \in G$
such that $f(x) = f(\sigma \cdot x)$ for all $x \in X$. We say $f$ and $g$ are
\emph{sort-equivalent} if they are $\sym_{A_1} \times \ldots
\sym_{A_p}$-equivalent.

We call a function of the form $F:\{0,1\}^{X} \rightarrow \{0,1\}$ a
\emph{$(\tau, A)$-structured function}. We call $(\tau, A)$ the \emph{type} of
$F$, $\tau$ the \emph{signature of $F$} and $A$ \emph{the universe of $F$}. We
also use $\ind(F)$ to abbreviate $\ind (\type (F))$.Call $F$ a \emph{structured
  function} if it is a $(\tau, A)$-structured function for some appropriate
$\tau$ and $A$.

@@

These structured functions give us a general framework for indexing the input to
functions. Moreover, the structure on these index sets allows us to develop a
natural notion of symmetry particular to the indexing used by the function. Let
$F$ be a $(\tau, A)$-structured function. We say $F$ is \emph{$G$-invariant} if
for all $G$-equivalent $\vec{x},\vec{y} \in \{0,1\}^{X}$, $F(\vec{x}) =
F(\vec{y})$. We say $F$ is \emph{sort-invariant} if it is $(\tau, \sym_{A_{1}}
\times \ldots \times \sym_{A_{s}})$-invariant. If $\tau = (\{R\}, [2], \nu)$,
with $\nu (R) = (1, 2)$ and $F$ is sort-invariant then we say $F$ is
\emph{matrix-invariant}.

@@

% If $\tau$ is single-sorted we call $F$ \emph{simple} and if it consists of a
% single unary relation, we call $F$ \emph{unary-relational}.
We have developed this more general notion of symmetry for Boolean functions
that naturally takes into account the structure on the input sequence. The
following lemma shows that symmetry is just the requirement that the function be
invariant under permutations of its input, irrespective of structure.

\begin{lem}
  Let $F$ be a $(\tau, A)$-structured function. $F$ is a symmetric function if
  and only if $F$ is $\sym_A$-invariant.
\end{lem}

@@

\subsection{Symmetric Circuits}
Having developed the notion of a function that accepts input structured in
accordance with some vocabulary, we now develop a circuit model which
incorporates this structure on the inputs of a gate. We use this general
framework to define the notion of a matrix-symmetric circuit, a circuit with
gates labelled by matrix-symmetric Boolean functions, and finally we develop the
notion of a matrix-symmetric circuit with rank.

@@

\begin{definition}[Circuits on Structures]
  Let $\mathbb{B}$ be a basis of structured functions and $\tau := (R, \{s_1,
  \ldots, s_p\}, \nu)$ be a many-sorted vocabulary and let $\vec{n} := (n_{s_1},
  \ldots , n_{s_p}) \in \mathbb{N}^{S}$, we define a \emph{$(\mathbb{B},
    \tau)$-circuit} $C_{\vec{n}}$ computing a $q$-ary query $Q$ of type
  $(s^Q_{1}, \ldots , s^Q_{q})$ is a structure $\langle G, W, \Omega, \Sigma,
  \Lambda, L\rangle$.
  \begin{itemize}
    \setlength\itemsep{0mm}
  \item $G$ is called the set of gates of $C_{\vec{n}}$ and $\vert C_{\vec{n}}
    \vert := \vert G \vert$.
  \item $W \subseteq G \times G$, where $W$ is called the wires of the circuit.
    $(G,W)$ must be a directed acyclic graph. For $g \in G$ we $H_g := \{ h \in
    G: W(h,g)\}$ be the set of children of $g$.
  \item $\Omega$ is an injective function from $[n_{i_1}] \times \ldots \times
    [n_{i_q}]$ to $G$. The gates in the image of $\Omega$ are called the output
    gates. When $q = 0$, $\Omega$ is a constant function mapping to a single
    output gate.
  \item $\Sigma$ is a function from $G$ to $\mathbb{B} \uplus \tau \uplus
    \{0,1\} $ which maps input gates to $\tau \uplus \{0,1\}$ and where $\vert
    \Sigma^{-1} (0) \vert \leq 1$ $\vert \Sigma^{-1} (1) \vert \leq 1$ and the
    internal gates get mapped into $\mathbb{B}$. Gates mapped to $R$ are called
    relational gates and gates mapped to 1 or 0 are called constant gates.
  \item $\Lambda$ is a sequence of injective functions $(\Lambda_{R'})_{R' \in
      R}$ where for each $R' \in R$ with arity $r$ and type $(s^{R'}_{1},
    \ldots, s^{R'}_{r})$, $\Lambda_{R'}$ maps each relational gate $g$ with $R'
    = \Sigma (g)$ to the tuple $\Lambda_{R'} (g) \in [n_{s^{R'}_1}] \times
    \ldots \times [n_{S^{R'}_r}]$. When no ambiguity arises we write $\Lambda
    (g)$ for $\Lambda_{R'} (g)$.
  \item $L$ maps to each internal gate a labelling on its children. Let $g$ be
    an internal gate. We have that $L(g)$ is a surjection from $\ind (\Sigma (g))$
    to $H_g$. We call $L(g)$ the \emph{child-labelling} for $g$.
  \end{itemize}
\end{definition}

We write $C_{\vec{n}}$ to emphasise that the circuit accepts input structures
with sorted-size $\vec{n}$. If $\tau$ is single-sorted vocabulary then we write
$C_n$ to emphasise that $C_n$ accepts input structures of size $n$. For a gate
$g \in C_{\vec{n}}$ we abuse notation and define the \emph{type} of $g$, denoted
by $\type (g)$, to be type of $\Sigma(g)$, and the \emph{index} of $g$, denoted by
$\ind(g)$, to be the the index for $\Sigma(g)$.

Let $\tau = (\{s_1, \ldots, s_p\}, \{R_1, \ldots , R_t\}, \nu)$. Let
$\mathcal{A}$ be a many-sorted $\tau$-structure with sorts $U_1, \ldots, U_p$
and sort-size $\vec{n}$, and let $\gamma: U \rightarrow [n_{s_1}] \sqcup \ldots
\sqcup [n_{s_p}]$ be a sorted bijection. Let $\gamma \mathcal{A}$ be the
structure formed by mapping the universe of $\mathcal{A}$ in accordance with
$\gamma$. The evaluation of some $(\mathbb{B}, \tau)$-circuit $C$ computing a
q-ary query $Q$ of type $(s^Q_{1}, \ldots , s^Q_{q})$ proceeds by recursively
evaluating gates. The evaluation of the gate $g$ using $\gamma$ with input
$\mathcal{A}$ is denoted by $C[\gamma \mathcal{A}](g)$. The evaluation of a gate
$g$ is given as follows.
\begin{enumerate}
  \setlength\itemsep{0mm}
\item If $g$ is a constant gate then it evaluates to the bit given by
  $\Sigma(g)$.
\item If $g$ is a relational gate then $g$ evaluates to true iff $\gamma
  \mathcal{A} \models \Sigma(g)(\Lambda (g))$.
\item If $g$ is an internal gate labelled by a $(\tau_g, A_g)$-structured
  function $\Sigma(g)$, let $L^{\gamma}_g: \ind(F) \rightarrow \{0,1\}$ be
  defined by $L^{\gamma}_g(x) = C[\gamma \mathcal{A}](L(g)(x))$. Then $g$
  evaluates to true iff $\Sigma(g) (L^{\gamma}_g)$.
\end{enumerate}
$C$ defines the $q$-ary query $Q \subseteq U_{s^Q_1} \times \ldots \times
U_{s^Q_q}$ where $\vec{a} \in Q$ iff $C[\gamma \mathcal{A}](\Omega (\gamma
\vec{a})) = 1$.

The following definition is an important circuit property introduced by Anderson
and Dawar \cite{AndersonD17}. We introduce it for the sake of comparison.

\begin{definition}[Invariant Circuit]
  Let $C_{\vec{n}}$ be a $(\mathbb{B}, \tau)$-circuit, computing some $q$-ary
  query of type $(s^Q_1, \ldots , s^Q_q)$. We say $C_{\vec{n}}$ is
  \emph{invariant} if for every $\tau$-structure $\mathcal{A}$ of size
  $\vec{n}$, $\vec{a} \in U_{i_1} \times \ldots \times U_{i_q}$, and sorted
  bijections $\gamma_1, \gamma_2: U \rightarrow [n_{s_p}] \sqcup\ldots \sqcup
  [n_{s_p}]$ we have that $C[\gamma_1 \mathcal{A}](\Omega (\gamma_1 \vec{a})) =
  C[\gamma_2 \mathcal{A}](\Omega (\gamma_2 \vec{a}))$.
\end{definition}

This property ensures that the function computed by a circuit (or a family of
circuits) is invariant under isomorphisms on the input structure. In other
words, the circuit (or family of circuits) decides a property of the structure.
The following lemma allows us to recast this notion in terms of the language
developed in this paper.

\begin{lem}
  Let $C_n$ be a $(\mathbb{B}, \tau)$-circuit. We have that the function
  computed by $C_n$ is $\tau$-symmetric iff $C_n$ is invariant.
\end{lem}

>>>>

\documentclass[../paper.tex]{subfiles}
%\usepackage{../mymacros}
\begin{document}
% Note: include circuits in background
Boolean circuits have long been as an alternative model from which to study
algorithmic complexity. A circuit is evaluated for some ordered binary input
sequence by recursively evaluating each gate in the circuit, starting from the
input gates. Suppose then we are evaluating some gate $g$ in a circuit $C$. The
procedure involves taking the input string formed by evaluating all gates input
to $g$ and applying the Boolean function that labels $g$ to that string. But
this process assumes an externally-imposed ordering on the gates input to $g$;
an assumption common in most parts of complexity theory, where the encoding of
structures for computation induces such orderings, but not in finite model
theory.

Circuit models have also been studied in finite model theory as a means for
characterising the expressive power of logics. In particular, research has
focused on symmetric circuits defined over bases of symmetric Boolean functions
(see \cite{}). The fact that the basis consist of only symmetric functions is a
crucial requirement, ensuring that the evaluation of a gate does not depend on
the order.

In this section we extend the circuit model so as to allow us to define
symmetric circuits over a broader basis of functions. We introduce the notion of
a structured function , i.e. a Boolean function whose input is indexed in
accordance with some many-sorted structure. This allows us to characterise many
natural symmetry properties of structured functions in terms of equally natural
symmetry properties of many-sorted structures. We use this to derive a general
symmetry conditions on structured functions and a symmetric circuit model whose
basis elements are symmetric structured functions.

% We incorporate this labelling in the circuit model by including for each gate
% $g$ labelled by a structured function $$an indexing on its children that
% matches the index for the function that labels $g$. We allow these circuits to
% be defined over bases of structured functions with this more general symmetry
% condition

% In order then to include a gate computing rank in our symmetric circuit model
% we need to rec We are interested in circuit with gates that can compute rank
% and so, since rank is not a symmetric function, we develop a symmetric circuit
% model capable of incorporating non-symmetric functions in its basis. Moreover,
% we note that

% In particular, we are interested in Boolean functions that compute rank, i.e.
% functions that take in a binary matrix and compute the rank of that matrix
% over some prime field and compare the result with some threshold. But clearly
% this rank function is not symmetric. Furthermore, its evaluation depends on
% how the input set is labelled to form a matrix.

% In this section we develop the

% allowing us to develop natural symmetry conditions on the functions in terms
% of symmetry conditions on the structure. This indexing allows us to generalise
% the notion of a symmetric function by relativising the symmetry condition in
% accord with the structure on the input string. Incorporating this indexing
% into the circuit and

% ore recently, they have played a role in finite model theory, with many
% authors (e.g. ) studying circuit models that take in some representation of a
% finite algebraic structure and which obey certain natural symmetry conditions.
% However, there are some subtleties in the definition and evaluation of
% circuits which must be considered carefully and have important implications
% for the notion of a symmetric circuit studied in this paper. Let $C$ be a
% Boolean circuit $g$ be an internal gate labelled by some function $f_g:
% \{0,1\}^{A} \rightarrow \{0,1\}$, for some index set $A$ with $\vert A \vert =
% \vert H_g \vert$. Formally, evaluating $g$ requires a (usually bijective)
% indexing function $f_i: A \mapsto H_g$, which organises the inputs of $g$ into
% an appropriate sequence. The evaluation of $g$ is then given by $C[\vec{x}](g)
% = f_g (a \mapsto C[\vec{x}](f_i(a)))$. As such the evaluation of a gate in a
% circuit may depend on how the inputs to the gate are indexed.

% In most contexts in complexity theory, circuits may be taken to be implicitly
% ordered objects and so this indexing function can be defined as taking the
% $i$th element of $A$ to the $i$th element of $H_g$. However, in general this
% renders the evaluation of a circuit for a given input a property both of the
% circuit and an arbitrary order. In order to address this problem this problem
% many authors only define circuits over Boolean bases of symmetric functions
% (e.g. the standard basis or the majority basis). In this case the evaluation
% of a gate is independent of the choice of indexing function.

% However, in this paper we require gates labelled by Boolean rank functions,
% i.e. a function that takes in a Boolean sequence indexed by `row' and `column'
% sets, outputs 1 iff that matrix, when interpreted as having entries in
% $\mathbb{F}_p$, has rank at least $r$. These rank functions clearly depend on
% how the input sequence is labelled, and so are not symmetric. However, it is
% worth noting that while such rank functions may be symmetric, they are
% symmetric in the weaker sense of being constant under permutations on the rows
% and columns of the input matrix.

% Motivated by the desire to study circuits with gates that compute rank, We
% develop a general framework for circuits that allows gates to be labelled by
% Boolean functions of the form $F: \{0,1\}^{X} \rightarrow \{0, 1\}$, where the
% labelling $X$ is the universe of a many-sorted structure over some vocabulary
% $\tau$. The Moreover, we generalise the notion of a circuit for structures
% (see Anderson and Dawar \cite{AndersonD17}), incorporate an appropriate
% labelling for the inputs of each gate and allowing for a Boolean basis
% including non-symmetric functions. Furthermore, the vocabulary $\tau$ will
% allow us to impose natural symmetry conditions on $F$, and so allowing us to
% develop a useful notion of a symmetric circuit.

% We also prove that no family symmetric functions defined over a basis of
% symmetric functions can improve on the expressive power of the circuit model
% of Anderson and Dawar \cite{AndersonD17}. Together with the main theorem of
% this paper this implies that these more general symmetric circuits are
% strictly more powerful

@@

\subsection{Structured Functions and Symmetry}
Before we begin this section let us review Boolean circuits and a few useful
symmetry conditions placed on these circuits. Let $X$ be a finite set and $F:
\{0,1\}^{X} \rightarrow \{0,1\}$ be a Boolean function. We call $X$ the
\emph{index} of $F$, and denote it by $\ind(F)$. It is often the case that the
index $X$ is in some sense structured, but let us first consider the case where
$X$ is an unstructured set. In this case the only natural symmetry condition to
consider is where $F$ is invariant under the action of $\sym_n$, i.e. $F$ is a
symmetric function. Suppose instead $X$ consists of all two element subsets of
the set $V$. Here we can think of $X$ as representing all potential edges in a
graph, and $\vec{x} \in \{0,1\}^X$ as defining a graph. It is natural then to
consider the case where $F$ determines a property of the graph, i.e. where $F$
is invariant under the action of $\sym_V$ on the input (i.e renaming of the
vertices). We call such a function \emph{graph-invariant}. Suppose that $X := A
\times B$ for two non-empty sets $A$ and $B$. Here we can think of $\vec{x} \in
\{0,1\}^X$ as an (unordered) matrix, and we can consider the case where $F$ is
invariant under the action of $\sym_A \times \sym_B$. In this case we say $F$ is
\emph{matrix-invariant}. Notice that in each of these cases the symmetry
condition of interest on the function was related to the structure of the index
set.

In the rest of this subsection we develop the notion of structured function, a
Boolean function whose input is structured in accord with some structure and
whose natural symmetry conditions follow from similar conditions on the
structure.

% $X:= {n}\choose{2}$, for some $n$, we can think of $F$ as taking in some $n$
% vertex graph, and we should like $F$ to respect permutations on the graph. In
% this case we insist that $F$ is a \emph{graph symmetric} function, for all $$
% where we can think of $F$ as taking in a graph with $n$ vertices, where $X:=
% {n}\choose{2}$ as indexing a graph with $n$ vertices

% The set $X$ is often used just to provide an ordering, and so $F$ just takes
% in a binary $\vert X \vert$-tuple. While in general the order of the input
% sequence matters, it is common to consider the case where $F$ is a
% \emph{symmetric} Boolean function, i.e. for all $\sigma \in \sym_{X}$,
% $\vec{x} \in \{0,1\}^{X}$, $F(\vec{x}\cdot \sigma) = \vec{x}$. In this case
% the output of the function is independent of the ordering on the input tuple.
% Indeed, many natural Boolean functions (e.g. `and', `or', `not', `majority',
% etc.) are symmetric functions. One of the benefits of considering symmetric
% functions

Let $\tau = (R, S, \nu)$ be a many-sorted, relational signature, with $R = \{R_1
\ldots, R_t\}$ and $S = [s]$ for some $s \in \nats$, and $\nu(R_i) = (a^i_1,
\ldots , a^i_{r_i)})$, where $r_i = \arty(R_i)$ and $a^i_q \in S$ for all $q \in
[r_i]$. Let $\mathcal{D}$ be a $\tau$-structure, with the universe $D = D_1
\uplus \ldots \uplus D_s = \{(i,d) : d \in D_i\}$.

Let the \emph{index-domain} of $\mathcal{D}$ be $\rdom {\mathcal{D}} = \{(i, j)
:= i \in [r_j], R_j \text{ a relation in $\tau$} \}$. Let the \emph{index
  defined by} $\mathcal{D}$ be the set of all functions $f: \rdom(\mathcal{D})
\rightarrow D$ such that for all $R_j \in R$ and $i \in [r_j]$, $f(i, j) \in
A_{a^{j}_i}$. We abuse notation and denote the index defined by
$\ind(\mathcal{D})$. We note that, since the set of relation symbols and sort
symbols are ordered, we can also think of each $f \in \ind(\mathcal{D})$ as a
tuple in $D_{a^1_1} \times \ldots \times D_{a^1_{r_1}} \times \ldots \times
D_{a^t_1} \times D_{a^t_{r_t}}$. We will use these tuples in order to index the
inputs to a Boolean function.

For the remainder of the section fix a $\tau$-structure $\mathcal{D}$ and let $X
:= \ind (\mathcal{D})$. Let $G \leq \sym_{X}$. We note that there is a natural
action of $G$ on $X$. Furthermore, we can define an action of $G$ on functions
of the form $f: X \rightarrow H$, for any set $H$, given by $\sigma \cdot f(x) =
f(\sigma \cdot x)$ for all $x \in X$. If $H = \{0,1\}$ we say that $f$ is a
\emph{binary sequence indexed by $X$}. If $H$ is surjective, we say that
\emph{$f$ indexes $H$ by $X$} or \emph{$f$ indexes $H$ by $\mathcal{D}$}. We
note that $\aut(\mathcal{D})$ and $\iso(\mathcal{D})$ can be identified with
subgroups of $\sym_X$.

We develop a notion of equivalence from this action. Say that two functions $f,
g : X \rightarrow H$ are \emph{$G$-equivalent} iff there exists $\sigma \in G$
such that $f(x) = f(\sigma \cdot x)$ for all $x \in X$. We say $f$ and $g$ are
\emph{isomorphism-equivalent} if they are $\iso(\mathcal{D}$-equivalent. We say
that $f$ and $g$ are \emph{automorphism-equivalent} if they are $
\aut(\mathcal{D})$-equivalent. We say that $f$ and $g$ are
\emph{sort-equivalent} in the special case where every relation in $\mathcal{D}$
contains all possible tuples and $f$ and $g$ are automorphism-equivalent.

We call a function of the form $F:\{0,1\}^{X} \rightarrow \{0,1\}$ a
\emph{$\mathcal{D}$-structured function}. We note that $\ind(F) =
\ind(\mathcal{D})$, justifying our above choice of notation. We also say
\emph{the vocabulary of $F$} and \emph{the universe of $F$} to refer to the
vocabulary and universe of $\mathcal{D}$.

@@

These structured functions give us a general framework for indexing the input to
functions while also providing us with a natural set of groups from which to
develop symmetry conditions for Boolean functions. We now introduce some of
these conditions. Let $F$ be a $\mathcal{D}$-structured function. We say $F$ is
\emph{$G$-invariant} if it constant on $G$-equivalent input vectors, i.e. for
all $G$-equivalent $\vec{x},\vec{y} \in \{0,1\}^{X}$, $F(\vec{x}) = F(\vec{y})$.
Similarly, we say $F$ is \emph{automorphism-invariant} if it is constant on
automorphism-equivalent input vectors and \emph{isomorphism-invariant} if it is
constant on isomorphism-equivalent input vectors. If all the relations in
$\mathcal{D}$ contain all possible tuples, and $F$ is automorphism-invariant,
then we say that $F$ is \emph{sort-invariant}. If the vocabulary of $F$ has two
sorts $s_1, s_2$ and one binary relation $R$ with type $(s_1, s_2)$ and $F$ is
sort-invariant, then we say $F$ is \emph{matrix-invariant}.

Notice that if $\tau$ is the empty vocabulary and $\mathcal{D}$ is a
$\tau$-structure, i.e. $\mathcal{D}$ is a pure set, then any
$\mathcal{D}$-structured function $F$ is automorphism-invariant (or
isomorphism-invariant) if, and only if, $F$ is symmetric. Also, if $\mathcal{D}$
is a graph, then $F$ is isomorphism invariant if, and only if, $F$ is
graph-invariant (except with a slight alteration, here $F$ takes in the edge
relation rather than a binary string indexed by two element sets of vertices).
Similarly, $F$ is matrix invariant (as a structured function) if, and only if,
it is matrix-invariant in the sense defined right at the beginning of this
subsection. In each of these cases choosing an appropriate structure leads to an
appropriate index and an appropriate notion of symmetry. In this way structured
functions generalise the notion of a symmetric function and allow us to capture
a broad range of structure-specific symmetry conditions in a single definition.
In the next section we will show how to define a circuit and a symmetric circuit
using bases of structured functions.

@@

% If $\tau$ is single-sorted we call $F$ \emph{simple} and if it consists of a
% single unary relation, we call $F$ \emph{unary-relational}.

@@

\subsection{Symmetric Circuits}
We define a circuit over a basis of structured functions that are not
necessarily symmetric. As such, the circuit must include for each gate $g$ an
index on its children corresponding to the index of the function labelling $g$.

% Having developed the notion of a function that accepts input structured in
% accordance with some vocabulary, we now develop a circuit model which
% incorporates this structure on the inputs of a gate. We use this general
% framework to define the notion of a matrix-symmetric circuit, a circuit with
% gates labelled by matrix-symmetric Boolean functions, and finally we develop
% the notion of a matrix-symmetric circuit with rank.

@@

\begin{definition}[Circuits on Structures]
  Let $\mathbb{B}$ be a basis and $\tau$ be a relational vocabulary, we define a
  \emph{$(\mathbb{B}, \tau)$-circuit} $C_n$ computing a $q$-ary query $Q$ as a
  structure $\langle G, \Omega, \Sigma, \Lambda, L \rangle$.
  \begin{itemize}
    \setlength\itemsep{0mm}
  \item $G$ is called the set of gates of $C_n$ and $\vert C_n \vert := \vert G
    \vert$.
  \item $\Omega$ is an injective function from $[n]^q$ to $G$. The gates in the
    image of $\Omega$ are called the output gates. When $q = 0$, $\Omega$ is a
    constant function mapping to a single output gate.
  \item $\Sigma$ is a function from $G$ to $\mathbb{B} \uplus \tau \uplus
    \{0,1\} $ such that $\vert \Sigma^{-1} (0) \vert \leq 1$ $\vert \Sigma^{-1}
    (1) \vert \leq 1$. Those gates mapped to $\tau \uplus \{0,1\}$ are called
    input gates, with those mapped to $\tau$ called relational gates and those
    mapped $\{0,1\}$ called constant gates. Those gates mapped to $\mathbb{B}$
    are called internal gates.
  \item $\Lambda$ is a sequence of injective functions $(\Lambda_{R_i})_{R_i \in
      R}$ such that $\Lambda_{R_i}$ maps each relational gate $g$ with $R_i =
    \Sigma (g)$ to the tuple $\Lambda_{R_i} (g) \in [n]^{r}$. When no ambiguity
    arises we write $\Lambda (g)$ for $\Lambda_{R_i} (g)$.
  \item $L$ associates with each internal gate $g$ a function $L(g):
    \dom(\Sigma(g)) \rightarrow G$ such that if we define a relation $W
    \subseteq G^{2}$ by $W(h_1,h_2)$ iff $h_2$ is an internal gate and $h_1$ is
    in the image of $L(h_2)$, then $(G, W)$ is a directed acyclic graph.
  \end{itemize}
\end{definition}

% \begin{definition}[Circuits on Structures]
%   Let $\mathbb{B}$ be a basis of structured functions and $\tau := (R, \{s_1,
%   \ldots, s_p\}, \nu)$ be a many-sorted vocabulary and let $\vec{n} :=
%   (n_{s_1}, \ldots , n_{s_p}) \in \mathbb{N}^{S}$, we define a
%   \emph{$(\mathbb{B}, \tau)$-circuit} $C_{\vec{n}}$ computing a $q$-ary query
%   $Q$ of type $(s^Q_{1}, \ldots , s^Q_{q})$ is a structure $\langle G, W,
%   \Omega, \Sigma, \Lambda, L\rangle$.
%   \begin{itemize}
%     \setlength\itemsep{0mm}
%   \item $G$ is called the set of gates of $C_{\vec{n}}$ and $\vert C_{\vec{n}}
%     \vert := \vert G \vert$.
%   \item $W \subseteq G \times G$, where $W$ is called the wires of the
%     circuit. $(G,W)$ must be a directed acyclic graph. For $g \in G$ we $H_g
%     := \{ h \in G: W(h,g)\}$ be the set of children of $g$.
%   \item $\Omega$ is an injective function from $[n_{i_1}] \times \ldots \times
%     [n_{i_q}]$ to $G$. The gates in the image of $\Omega$ are called the
%     output gates. When $q = 0$, $\Omega$ is a constant function mapping to a
%     single output gate.
%   \item $\Sigma$ is a function from $G$ to $\mathbb{B} \uplus \tau \uplus
%     \{0,1\} $ which maps input gates to $\tau \uplus \{0,1\}$ and where $\vert
%     \Sigma^{-1} (0) \vert \leq 1$ $\vert \Sigma^{-1} (1) \vert \leq 1$ and the
%     internal gates get mapped into $\mathbb{B}$. Gates mapped to $R$ are
%     called relational gates and gates mapped to 1 or 0 are called constant
%     gates.
%   \item $\Lambda$ is a sequence of injective functions $(\Lambda_{R'})_{R' \in
%     R}$ where for each $R' \in R$ with arity $r$ and type $(s^{R'}_{1},
%     \ldots, s^{R'}_{r})$, $\Lambda_{R'}$ maps each relational gate $g$ with
%     $R' = \Sigma (g)$ to the tuple $\Lambda_{R'} (g) \in [n_{s^{R'}_1}] \times
%     \ldots \times [n_{S^{R'}_r}]$. When no ambiguity arises we write $\Lambda
%     (g)$ for $\Lambda_{R'} (g)$.
%   \item $L$ maps to each internal gate a labelling on its children. Let $g$ be
%     an internal gate. We have that $L(g)$ is a surjection from $\ind (\Sigma
%     (g))$ to $H_g$. We call $L(g)$ the \emph{child-labelling} for $g$.
%   \end{itemize}
% \end{definition}

Let $C_n = \langle G, \Omega, \Sigma, \Lambda, L \rangle$ be a $(\mathbb{B},
\tau)$-circuit. The $n$ here denotes that $C_n$ accepts structures of size $n$.
For a gate $g \in C_n$, we define the \emph{index} of $g$, denoted by $\ind(g)$,
as $\ind(\Sigma(g))$ and we let $H_g$ denote the set of all gates in the image
of $L(g)$.

For any two sets $X, Y$ we let $X^{\underline{Y}}$ denote the set of all
injective functions from $Y$ to $X$.

Let $\tau = (R_1, \ldots , R_t)$, $\mathcal{A}$ be a $\tau$-structure of size
$n$, and $\gamma: \in [n]^{\underline{U}}$. Let $\gamma \mathcal{A}$ be the
structure formed by mapping U, the universe of $\mathcal{A}$, in accordance with
$\gamma$. The evaluation of some $(\mathbb{B}, \tau)$-circuit $C$ computing a
q-ary query $Q$ proceeds by recursively evaluating the gates in the circuit. The
evaluation of the gate $g$ using $\gamma$ with input $\mathcal{A}$ is denoted by
$C[\gamma \mathcal{A}](g)$, and is given as follows
\begin{enumerate}
  \setlength\itemsep{0mm}
\item If $g$ is a constant gate then it evaluates to the bit given by
  $\Sigma(g)$.
\item If $g$ is a relational gate then $g$ evaluates to true iff $\gamma
  \mathcal{A} \models \Sigma(g)(\Lambda (g))$.
\item If $g$ is an internal gate such that $\Sigma (g)$ is a
  $\mathcal{D}$-structured function, let $L^{\gamma}_g: \ind(g) \rightarrow
  \{0,1\}$ be defined by $L^{\gamma}_g(x) = C[\gamma \mathcal{A}](L(g)(x))$, for
  all $x \in \ind(g)$. Then $g$ evaluates to true iff $\Sigma(g) (L^{\gamma}_g)
  = 1$.
\end{enumerate}
We say that $C_n$ defines the $q$-ary query $Q \subseteq U^q$ where $\vec{a} \in
Q$ iff $C_n[\gamma \mathcal{A}](\Omega (\gamma \vec{a})) = 1$.

% The circuit $C_n$ The following definition is an important circuit property
% introduced by Anderson and Dawar \cite{AndersonD17}. We introduce it for the
% sake of comparison.

The following circuit condition is from Anderson and Dawar \cite{AndersonD17}.

\begin{definition}[Invariant Circuit]
  Let $C_n$ be a $(\mathbb{B}, \tau)$-circuit, computing some $q$-ary query. We
  say $C_n$ is \emph{invariant} if for every $\tau$-structure $\mathcal{A}$ of
  size $n$, $\vec{a} \in U^{q}$, and $\gamma_1, \gamma_2: [n]^{\underline{U}}$
  we have that $C[\gamma_1 \mathcal{A}](\Omega (\gamma_1 \vec{a})) = C[\gamma_2
  \mathcal{A}](\Omega (\gamma_2 \vec{a}))$.
\end{definition}
In general, the evaluation of a circuit depends on the presentation of the
structure (i.e. on the chosen bijection $\gamma$). If family of $(\mathcal{B},
\tau)$-circuits $\mathcal{C}$ is invariant it follows that the query computed is
a $q$-query on $\tau$-structures. It follows that if $q = 0$ then $\mathcal{C}$
computes a property of $\tau$-structures. The following lemma allows us to
recast this notion in terms of the language developed in this paper.
